DNA Methylation Analyses

### 2023-02-17 DNA-methylation.Rmd
fmt

# Introduction
The purpose of this tutorial is to provide an example of taking Bisulfite treated DNA sequence data through identification of methylation at the loci and gene level. The principle behind creating bisulfite treated DNA libraries is that when cytosines are not methylated they will be converted to uracil, which ultimately result in thymine. When aligning to a reference genome the percent methylation at a given CpG loci is determined by examine the ratio to thymines and cytosines. For example if there is a given CpG loci with 30% of the alignment with the cytosine on the reference being and thymine, one would consider this cytosine loci (CpG) to be 70% methylated.

@@

# Sequence Quality
An example of what raw data might look like from a marine invertebrate is shown. _Note that DNA methylation patters are diverse across taxa with most marine invertebrates demonstrating a mosiac pattern and global methylation at CpG loci is approximately 15%._

@@

```
/gscratch/srlab/programs/TrimGalore-0.4.5/trim_galore \
--output_dir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD \
--paired \
--fastqc_args \
--outdir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD/FASTQC \
--threads 28 \
--illumina \
--clip_R1 10 \
--clip_R2 10 \
--three_prime_clip_R1 10 \
--three_prime_clip_R2 10 \
--path_to_cutadapt /gscratch/srlab/programs/miniconda3/bin/cutadapt \
/gscratch/scrubbed/sr320/froger-raw/00_fastq/Meth17_R1_001.fastq.gz \
```

@@

# Read Alignment
The software used as part of this tutorial is **Bismark Bisulfite Mapper**. As with any software it is best to be famililar with the [User Manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html). Thus content below will provided with the assumption that the reader has read the manual and is meant to serve as guidance based on experience working with marine invertebrates.

@@

```
# Directories and programs
bismark_dir="/programs/Bismark-0.21.0"
bowtie2_dir="/programs/bowtie2-2.3.4.1-linux-x86_64/"
genome_folder="/where/the/fastafile/lives/"

@@

```

@@

```
./Bisulfite_Genome
./Bisulfite_Genome/GA_conversion
./Bisulfite_Genome/CT_conversion
```

@@

```
find ${reads_dir}*_R1_001_val_1.fq.gz \
| xargs basename -s _R1_001_val_1.fq.gz | xargs -I{} ${bismark_dir}/bismark \
--path_to_bowtie ${bowtie2_dir} \
-genome ${genome_folder} \
-p 4 \
-score_min L,0,-0.6 \
--non_directional \
-1 ${reads_dir}{}_R1_001_val_1.fq.gz \
-2 ${reads_dir}{}_R2_001_val_2.fq.gz \
-o Mcap_tg
```

@@

```
find *.bam | \
xargs basename -s .bam | \
xargs -I{} ${bismark_dir}/deduplicate_bismark \
--bam \
--paired \
{}.bam
```

# Methylation Quantification
Methylation levels are then extracted using the `bismark_methylation_extractor`. For example:

```
${bismark_dir}/bismark_methylation_extractor \
--bedGraph \
--counts \
--comprehensive \
--merge_non_CpG \
--multicore 28 \
--buffer_size 75% \
*deduplicated.bam
```

@@

```
NC_035784.1 141 141 37.5    3   5
NC_035784.1 142 142 100 2   0
NC_035784.1 155 155 70  7   3
NC_035784.1 156 156 100 2   0
NC_035784.1 291 291 0   0   2
NC_035784.1 292 292 0   0   3
NC_035784.1 313 313 0   0   1
NC_035784.1 314 314 66.6666666666667    2   1
NC_035784.1 470 470 66.6666666666667    4   2
NC_035784.1 611 611 0   0   4
```

@@

```
<chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>
```

@@

```
find *deduplicated.bismark.cov.gz \
| xargs basename -s _trimmed_bismark_bt2.deduplicated.bismark.cov.gz \
| xargs -I{} ${bismark_dir}/coverage2cytosine \
--genome_folder ${genome_folder} \
-o {} \
--merge_CpG \
--zero_based \
{}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz
```

@@

```
NC_035785.1 217 219 100.000000  17  0
NC_035785.1 523 525 87.500000   7   1
NC_035785.1 556 558 50.000000   5   5
NC_035785.1 727 729 100.000000  16  0
NC_035785.1 1330    1332    0.000000    0   2
NC_035785.1 1403    1405    0.000000    0   2
NC_035785.1 1494    1496    66.666667   2   1
NC_035785.1 1747    1749    100.000000  8   0
NC_035785.1 2024    2026    100.000000  24  0
NC_035785.1 2054    2056    93.333333   14  1
```

@@

```
#creating bedgraphs post merge
for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
  > "${STEM}"_10x.bedgraph
done

@@

for f in *merged_CpG_evidence.cov
do
  STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
  cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4, $5, $6}}' \
  > "${STEM}"_5x.tab
done
```

@@

```
# Sort files for methylkit and IGV
find *.bam | \
xargs basename -s .bam | \
xargs -I{} ${samtools} \
sort --threads 28 {}.bam \
-o {}.sorted.bam

# Index sorted files for IGV
find *.sorted.bam | \
xargs basename -s .sorted.bam | \
xargs -I{} ${samtools} \
index -@ 28 {}.sorted.bam
```

>>>>

# Introduction

The purpose of this tutorial is to provide an example of taking Bisulfite treated DNA sequence data through identification of methylation at the loci and gene level. The principle behind creating bisulfite treated DNA libraries is that when cytosines are not methylated they will be converted to uracil, which ultimately result in thymine. When aligning to a reference genome the percent methylation at a given CpG loci is determined by examine the ratio to thymines and cytosines. For example if there is a given CpG loci with 30% of the alignment with the cytosine on the reference being and thymine, one would consider this cytosine loci (CpG) to be 70% methylated.

@@

# Sequence Quality

An example of what raw data might look like from a marine invertebrate is shown. *Note that DNA methylation patters are diverse across taxa with most marine invertebrates demonstrating a mosiac pattern and global methylation at CpG loci is approximately 15%.*

@@

    /gscratch/srlab/programs/TrimGalore-0.4.5/trim_galore \
    --output_dir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD \
    --paired \
    --fastqc_args \
    --outdir /gscratch/scrubbed/strigg/analyses/20200311/WGBS_MBD/FASTQC \
    --threads 28 \
    --illumina \
    --clip_R1 10 \
    --clip_R2 10 \
    --three_prime_clip_R1 10 \
    --three_prime_clip_R2 10 \
    --path_to_cutadapt /gscratch/srlab/programs/miniconda3/bin/cutadapt \
    /gscratch/scrubbed/sr320/froger-raw/00_fastq/Meth17_R1_001.fastq.gz \

@@

# Read Alignment

The software used as part of this tutorial is **Bismark Bisulfite Mapper**. As with any software it is best to be famililar with the [User Manual](https://rawgit.com/FelixKrueger/Bismark/master/Docs/Bismark_User_Guide.html). Thus content below will provided with the assumption that the reader has read the manual and is meant to serve as guidance based on experience working with marine invertebrates.

@@

    # Directories and programs
    bismark_dir="/programs/Bismark-0.21.0"
    bowtie2_dir="/programs/bowtie2-2.3.4.1-linux-x86_64/"
    genome_folder="/where/the/fastafile/lives/"

@@

@@

    ./Bisulfite_Genome
    ./Bisulfite_Genome/GA_conversion
    ./Bisulfite_Genome/CT_conversion

@@

    find ${reads_dir}*_R1_001_val_1.fq.gz \
    | xargs basename -s _R1_001_val_1.fq.gz | xargs -I{} ${bismark_dir}/bismark \
    --path_to_bowtie ${bowtie2_dir} \
    -genome ${genome_folder} \
    -p 4 \
    -score_min L,0,-0.6 \
    --non_directional \
    -1 ${reads_dir}{}_R1_001_val_1.fq.gz \
    -2 ${reads_dir}{}_R2_001_val_2.fq.gz \
    -o Mcap_tg

@@

    find *.bam | \
    xargs basename -s .bam | \
    xargs -I{} ${bismark_dir}/deduplicate_bismark \
    --bam \
    --paired \
    {}.bam

# Methylation Quantification

Methylation levels are then extracted using the `bismark_methylation_extractor`. For example:

    ${bismark_dir}/bismark_methylation_extractor \
    --bedGraph \
    --counts \
    --comprehensive \
    --merge_non_CpG \
    --multicore 28 \
    --buffer_size 75% \
    *deduplicated.bam

@@

    NC_035784.1 141 141 37.5    3   5
    NC_035784.1 142 142 100 2   0
    NC_035784.1 155 155 70  7   3
    NC_035784.1 156 156 100 2   0
    NC_035784.1 291 291 0   0   2
    NC_035784.1 292 292 0   0   3
    NC_035784.1 313 313 0   0   1
    NC_035784.1 314 314 66.6666666666667    2   1
    NC_035784.1 470 470 66.6666666666667    4   2
    NC_035784.1 611 611 0   0   4

@@

    <chromosome> <start position> <end position> <methylation percentage> <count methylated> <count unmethylated>

@@

    find *deduplicated.bismark.cov.gz \
    | xargs basename -s _trimmed_bismark_bt2.deduplicated.bismark.cov.gz \
    | xargs -I{} ${bismark_dir}/coverage2cytosine \
    --genome_folder ${genome_folder} \
    -o {} \
    --merge_CpG \
    --zero_based \
    {}_trimmed_bismark_bt2.deduplicated.bismark.cov.gz

@@

    NC_035785.1 217 219 100.000000  17  0
    NC_035785.1 523 525 87.500000   7   1
    NC_035785.1 556 558 50.000000   5   5
    NC_035785.1 727 729 100.000000  16  0
    NC_035785.1 1330    1332    0.000000    0   2
    NC_035785.1 1403    1405    0.000000    0   2
    NC_035785.1 1494    1496    66.666667   2   1
    NC_035785.1 1747    1749    100.000000  8   0
    NC_035785.1 2024    2026    100.000000  24  0
    NC_035785.1 2054    2056    93.333333   14  1

@@

    #creating bedgraphs post merge
    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 10) {print $1, $2, $3, $4}}' \
      > "${STEM}"_10x.bedgraph
    done

@@

    for f in *merged_CpG_evidence.cov
    do
      STEM=$(basename "${f}" .CpG_report.merged_CpG_evidence.cov)
      cat "${f}" | awk -F $'\t' 'BEGIN {OFS = FS} {if ($5+$6 >= 5) {print $1, $2, $3, $4, $5, $6}}' \
      > "${STEM}"_5x.tab
    done

@@

    # Sort files for methylkit and IGV
    find *.bam | \
    xargs basename -s .bam | \
    xargs -I{} ${samtools} \
    sort --threads 28 {}.bam \
    -o {}.sorted.bam

    # Index sorted files for IGV
    find *.sorted.bam | \
    xargs basename -s .sorted.bam | \
    xargs -I{} ${samtools} \
    index -@ 28 {}.sorted.bam
